### 2016-06-20 store-consistency.tex
Store Consistency.

\paragraph{An SC store} An SC store guarantees total order of all
operations w.r.t $\visZ$ that is consistent with their chronological
order. A straightforward $\I_s$ for this store formalizes this
guarantee:
\begin{smathpar}
\begin{array}{l}
  \I_s \;=\; \lambda\E.~\forall\eta_1,\eta_2.~\{\eta_1,\eta_2\}
  \subseteq \E.\A \conj \id(\eta_1) < \id(\eta_2) \\
  \hspace*{2in}\Rightarrow \underE{\eta_1 \visar \eta_2}
\end{array}
\end{smathpar}
Unfortunately, the above $\I_s$ is inconsistent with all isolation
specifications of Fig.~\ref{fig:ansi-isolation}. For instance, let
$\eta_1$ be an effect of an uncommitted transaction and let $\eta_2$
be an effect of a \iso{Read Committed} transaction such that
$\id(\eta_1) < \id(\eta_2)$. While \C{RC} prohibits $\eta_1$ from
being visible $\eta_2$, $\I_s$ requires $\eta_1$ to be visible to
$\eta_2$. The machine, which cannot satisfy both constraints
simultaneously, gets stuck.

The way out of this impasse is to recognize that SC property of the
store needs to be weakened to admit weak isolation.

>>>>

\paragraph{An SC store} An SC store guarantees total order of all
operations w.r.t $\visZ$ that is consistent with their chronological
order. A straightforward $\I_s$ for this store is the SC guarantee
defined below:
\begin{smathpar}
\begin{array}{l}
  \C{SC}(\E) \;=\; \forall\eta_1,\eta_2.~\{\eta_1,\eta_2\}
  \subseteq \E.\A \conj \id(\eta_1) < \id(\eta_2) \\
  \hspace*{2in}\Rightarrow \underE{\eta_1 \visar \eta_2}
\end{array}
\end{smathpar}
Unfortunately, $\I_s=\C{SC}$ conflicts with all isolation
specifications of Fig.~\ref{fig:ansi-isolation}. For instance,
consider a case where $\I_c(\E) \;=\; \forall
T_i.~\underE{\C{RC}(T_i)}$. That is, all transactions execute under
\iso{Read Committed} isolation. Imagine a sample execution where
$\eta_1$'s transaction is not yet committed when $\eta_2$ is
generated. Letting $\eta_1$ be visible to $\eta_2$ violates $\I_c$,
whereas not letting it be visible violates $\I_{s}$. The only way to
satisfy both the invariants is to rule out all the executions that
interleave the operations of one transaction with the other, thereby
enforcing serializability and the ACID model of
transactions\footnote{Thus, serializability is a natural
generalization of SC to transactions.}. In general, when $\I_s$ is in
conflict (but not inconsistent) with $\I_c$, the only way to enforce
both is to restrict concurrency. Clearly, this is unacceptable since
it defeats the very purpose of weak isolation. 
% How then do we enforce weak isolation on a strongly consistent
% machine?

Relational databases demonstrate a way of out this impasse.
Implementations of weak isolation levels on relational databases
implicitly relax the SC requirement so as to maximize concurrency and
improve performance. Consequently, execution traces do not necessarily
satisfy \C{SC}. This approach hints at a general solution reconciling
$\I_s$ with $\I_c$, which is to weaken the former. However, there are
many ways $\I_s$ can be weakened; for instance, by setting it to
$true$, which makes the store EC. In reality though, weak isolation on
databases does not cause so drastic a reduction in consistency. It is
therefore possible that there exists a a principled approach to weaken
$\I_s$ that adequately captures the underlying reality. Such a
principle can be gleaned by observing that data stores often make
\emph{recency} commitments~\cite{bailishat}, by which they aim to make
the most recent data available to clients. SC is infact an extreme
form of a recency commitment, which guarantees that reads witness most
recent writes to a data item. In case this conflicts with weak
isolation guarantees, such as RR which requires all reads in a
transaction to witness the same state, the natural way to weaken
consistency while prioritizing recency is to let the reads witness
most recent state that \emph{does not} violate the isolation
constraints. In the context of the operational model of
Fig.~\ref{fig:txnimp}, this translates to making visible to an
operation the largest subset ($S$) of the global state ($\A$) that
does not lead to the violation of isolation constraints. Generalizing
this approach to any $\I_s$ and $\I_c$ gives the \emph{maximum
visiblity principle}, which requires the weakened consistency
guarantee ($\I_s'$) of a store to enforce all visibility relationships
imposed by the actual consistency guarantee ($\I_s$) on a given trace,
unless enforcing such a relationship violates $\I_c$.  Formally,
maximum visibility weakening of $\I_s$ is an $\I_s'$, such that:
\begin{itemize}
  \item $\I_s'$ is weaker than $\I_s$: 
      $\forall\E.~ \I_s(\E) \Rightarrow \I_s'(\E)$, and
  \item In every trace $\E$ that satisfies $\I_c$, and for every pair
  of effects $\eta_1$ and $\eta_2$ in $\E$, if $\I_s(\E)$ requires
  $\eta_1$ to be visible to $\eta_2$, then so does $\I_s'(\E)$ unless
  extending $\E$ with $\visZ(\eta_1,\eta_2)$ violates
  $\I_c$\footnote{\GK{ToDo: consider other well-formedness conditions
  on trace, such as acyclicity of $\visZ$ and $\soZ$. Are they needed
  (considering that the machine never violates them)? Encode the
  specifications in Z3 and make sure they are consistent.}}:
  \begin{smathpar}
  \begin{array}{l}
  \forall\E,\eta_1,\eta_2.~ \I_c(\E) \Rightarrow (\I_s(\E)
    \Rightarrow \underE{\eta_1 \visar \eta_2}) \Rightarrow \\
    \hspace*{0.5in}(\I_s'(\E) \Rightarrow \underE{\eta_1 \visar
    \eta_2} \disj \neg\I_c(\E\,\cup\,(\emptyset,\{(\eta_1,\eta_2)\})))
  \end{array}
  \end{smathpar}
\end{itemize}

Applying maximum visibility principle, we weaken \C{SC} to obtain the
following store trace invariant ($\I_s$) for an SC store whose
isolation constraints are captured by $\I_c$:
\begin{smathpar}
\begin{array}{lcl}
\I_s(\E) & = & \forall \eta_1,\eta_2.\, \{\eta_1,\eta_2\},
    \subseteq \E.\A \conj \id(\eta_1) <
    \id(\eta_2) \\
    & & \hspace*{0.5in} \Rightarrow 
      \underE{\eta_1 \visar \eta_2} \disj \neg\I_c(\E
    \cup (\emptyset,\{(\eta_1,\eta_2)\}))\\
\end{array}
\end{smathpar}
As usual $\I = \I_s \wedge \I_c$. The resultant operational model
hides an existing effect ($\eta\in\A$) from being visible to the
current operation if and only if showing it results in the violation
of $\I_c$.

\paragraph{A CC store} A causally consistent data store, such as the
one underlying the reasoning framework of~\cite{gotsmanpopl16}, allows
operations to only witness a causally consistent snapshot of the
global state. A straightforward trace invariant ($\I_s$) for this
store is the \C{CC} guarantee that requires an effect to witness all
causally preceding effects regardless of the isolation requirements of
its transaction:
\begin{smathpar}
\begin{array}{lccl}
\C{CC}(\E) & \;=\; &  & \forall \eta_1,\eta_2.\, 
      \E \Vdash \eta_1 \soar \eta_2 \Rightarrow  \underE{\eta_1 \visar
      \eta_2}\\
    &   & \wedge & \forall\eta_1,\eta_2,\eta_3.\,\underE{\eta_1 \visar
      \eta_2} \conj \underE{\eta_2 \visoar \eta_3} \\
    &   & &\hspace*{0.5in} \Rightarrow \underE{\eta_1 \visar \eta_3}\\
\end{array}
\end{smathpar}
However, like $\I_s = \C{SC}$, $\I_s = \C{CC}$ may induce the store to
meet isolation requirements by restricting concurrency, thus failing
to abide by the spirit of weak isolation. In order to admit weak
isolation without restricting concurrency, we weaken $\I_s$ while
being guided by the maximum visibility principle. The weakened $\I_s$
is shown below:
\begin{smathpar}
\begin{array}{lccl}
\C{CC}(\E) & \;=\; &  & \forall \eta_1,\eta_2.\, 
      \E \Vdash \eta_1 \soar \eta_2 \Rightarrow  \underE{\eta_1 \visar
      \eta_2} \\
    & & & \hspace*{0.6in}\disj \neg\I_c(\E \cup 
                (\emptyset,\{(\eta_1,\eta_2)\}))\\
    &   & \wedge & \forall\eta_1,\eta_2,\eta_3.\,\underE{\eta_1 \visar
      \eta_2} \conj \underE{\eta_2 \visoar \eta_3} \\
    &   & &\hspace*{0.3in} \Rightarrow \underE{\eta_1 \visar \eta_3}
      \disj \neg\I_c(\E \cup (\emptyset,\{(\eta_1,\eta_2)\}))\\
\end{array}
\end{smathpar}
Instantiating the parameter $\I$  with $\I_s \wedge \I_c$ results in
an operational semantics that admits the violation of causal
consistency if and only if the violation is inevitable to enforce
$\I_c$.
